A Mechanical Formalization of Higher-Ranked Polymorphic Type Inference

### 2019-06-25 sections/algorithm.tex
minor format: capitalizations, etc
camera-ready: acmart ver., fig. Description

\framebox{$\text{wf }\Gm$} Well-formed algorithmic worklist
\begin{gather*}
\inferrule*[right=$\mathtt{wf\nil}$]
{~}{\text{wf }\nil}
\qquad
\inferrule*[right=$\mathtt{wf_a}$]
{\text{wf }\Gm}
{\text{wf }\Gm, a}
\qquad
\inferrule*[right=$\mathtt{wf_{\al}}$]
{\text{wf }\Gm}
{\text{wf }\Gm, \al}
\qquad
\inferrule*[right=$\mathtt{wf_{of}}$]
{\text{wf }\Gm \\ \Gm\vdash A}
{\text{wf }\Gm, x:A}
\qquad
\inferrule*[right=$\mathtt{wf_{\jg}}$]
{\text{wf }\Gm \\ \Gm\vdash\jg}
{\text{wf }\Gm \Vdash\jg}
\end{gather*}
\caption{Extended Syntax and Well-Formedness for the Algorithmic System}\label{fig:alg:syntax}
\end{figure}

@@

% TODO rules too long
\begin{figure}[htp]
\hfill \framebox{$\Gm\rto \Gm'$} \hfill $\Gm$ reduces to $\Gm'$.
%\jimmy{Whenever there is a change on judgment count, marker should be properly added/removed. Decidability: all valid judgment chain should be reduced to nil.}
\begin{gather*}
\begin{aligned}
\Gm, a &\algrule \Gm \qquad
\Gm, \al \algrule \Gm \qquad
\Gm, x:A \algrule \Gm
\\[3mm]
\Gm \Vdash 1\le 1 &\algrule \Gm\\
\Gm \Vdash a\le a &\algrule \Gm\\
\Gm \Vdash \al\le \al &\algrule \Gm\\
\Gm \Vdash A_1\to A_2 \le B_1\to B_2 &\algrule \Gm \Vdash A_2 \le B_2 \Vdash B_1\le A_1\\
\Gm \Vdash \all A\le B &\algrule \Gm,\al \Vdash [\al/a]A\le B \quad\text{when } B \neq \all B'\\
\Gm \Vdash A\le \all[b]B &\algrule \Gm,b \Vdash A\le B
\\[3mm]
%\\
%\text{Let } \color{red}\Gm[\al \toto B // G_M] &:= \Gm_L, G_M, [B/\al]\Gm_R, \text{ when } \Gm[\al] = \Gm_L,\al,\Gm_R \ (G_M\text{ defaults to }\nil)\\
\Gm[\al] \Vdash \al \le A\to B &\algrule [\al[1]\to\al[2]/\al] (\Gm[\al[1], \al[2]] \Vdash \al[1]\to \al[2] \le A \to B)\\
 &\qquad\qquad \text{when }\al\notin FV(A)\cup FV(B)\\
\Gm[\al] \Vdash A\to B \le \al &\algrule [\al[1]\to \al[2]/\al] (\Gm[\al[1], \al[2]] \Vdash A \to B \le \al[1]\to \al[2])\\
 &\qquad\qquad \text{when }\al\notin FV(A)\cup FV(B)
 \\[3mm]
\Gm[\al][\bt] \Vdash \al \le \bt &\algrule [\al/\bt](\Gm[\al][])\\
\Gm[\al][\bt] \Vdash \bt \le \al &\algrule [\al/\bt](\Gm[\al][])\\
\Gm[a][\bt] \Vdash a \le \bt &\algrule [a/\bt](\Gm[a][])\\
\Gm[a][\bt] \Vdash \bt \le a &\algrule [a/\bt](\Gm[a][])\\
\Gm[\bt] \Vdash 1 \le \bt &\algrule [1/\bt](\Gm[])\\
\Gm[\bt] \Vdash \bt \le 1 &\algrule [1/\bt](\Gm[])
\\[3mm]
\Gm \Vdash e \Lto B &\algrule \Gm \Vdash e\To_a a\le B \quad
    \text{when } e \neq \lam e' \text{ and } B \neq \all B'\\
% \Gm \Vdash () \Lto 1 &\algrule \Gm\\
\Gm \Vdash e\Lto \all A &\algrule \Gm,a \Vdash e\Lto A\\
\Gm \Vdash \lam e \Lto A\to B &\algrule \Gm, x:A  \Vdash e \Lto B\\
\Gm[\al] \Vdash \lam e \Lto \al &\algrule [\al[1]\to \al[2] / \al](\Gm[\al[1],\al[2]], x:\al[1] \Vdash e \Lto \al[2])
% \quad\text{\jimmy{Additional}}
\\[3mm]
\Gm \Vdash x\To_a \jg &\algrule \Gm \Vdash [A/a] \jg \quad \text{when } (x:A)\in \Gm\\
\Gm \Vdash (e:A)\To_a \jg &\algrule \Gm \Vdash [A/a]\jg \Vdash e \Lto A\\
\Gm \Vdash ()\To_a \jg &\algrule \Gm \Vdash [1/a]\jg\\
\Gm \Vdash \lam e \To_a \jg &\algrule
    \Gm,\al,\bt \Vdash [\al\to\bt/a]\jg, x:\al \Vdash e\Lto \bt\\
\Gm \Vdash e_1\ e_2 \To_a \jg &\algrule \Gm \Vdash e_1\To_b (\appInfAlg{b}{e_2})
\\[3mm]
\Gm \Vdash \appInfAlg{\all A}{e} &\algrule \Gm,\al \Vdash \appInfAlg{[\al/a]A}{e}\\
\Gm \Vdash \appInfAlg{A\to C}{e} &\algrule \Gm \Vdash [C/a]\jg \Vdash e \Lto A\\
\Gm[\al] \Vdash \appInfAlg{\al}{e} &\algrule
    [\al[1]\to\al[2]/\al](\Gm[\al[1], \al[2]] \Vdash \appInfAlg{\al[1]\to\al[2]}{e})
%	[\al[1]\to\al[2]/\al](\Gm[\al[1], \al[2]] \Vdash [\al[2]/a]\jg \Vdash e\Lto \al[1])\\
% &\color{magenta} \makebox[0pt]{\qquad or} \phantom{{}\rrule{}{}}
\end{aligned}
\end{gather*}
\caption{Algorithmic Typing}\label{fig:alg}
\end{figure}

@@

\paragraph{{\bf 1. Algorithmic Subtyping}}
We have six subtyping rules (Rules 4-9) that are similar to their
declarative counterparts. For instance, Rule 7 consumes a subtyping
judgment and pushes two back to the worklist.  Rule 8 differs from
declarative Rule $\mathtt{{\le}{\forall}L}$ by introducing an existential
variable $\al$ instead of guessing the monotype $\tau$
instantiation. Each existential variable is later solved to a
monotype $\tau$ with the same context, so the final solution $\tau$ of
$\al$ should be well-formed under $\Gm$.

@@

\begin{figure}
\begin{gather*}
\begin{aligned}
           & \nil \Vdash (\lam x)~() \Lto 1\\
\rrule{18} & \nil \Vdash (\lam x)~() \To_a a\le 1\\
\rrule{26} & \nil \Vdash (\lam x) \To_b (\appInfAlg{b}{()}[a][a\le 1])\\
\rrule{25} & \nil,\al,\bt \Vdash \appInfAlg{\al\to\bt}{()}[a][a\le 1], x:\al \Vdash x\Lto \bt\\
\rrule{18} & \nil,\al,\bt \Vdash \appInfAlg{\al\to\bt}{()}[a][a\le 1], x:\al \Vdash x\To_b b\le \bt\\
\rrule{22} & \nil,\al,\bt \Vdash \appInfAlg{\al\to\bt}{()}[a][a\le 1], x:\al \Vdash \al\le \bt\\
\rrule{12} & \nil,\al \Vdash \appInfAlg{\al\to\al}{()}[a][a\le 1], x:\al\\
\rrule{3}  & \nil,\al \Vdash \appInfAlg{\al\to\al}{()}[a][a\le 1]\\
\rrule{28} & \nil,\al \Vdash \al\le 1 \Vdash () \Lto \al\\
\rrule{18} & \nil,\al \Vdash \al\le 1 \Vdash () \To_a a\le \al\\
\rrule{24} & \nil,\al \Vdash \al\le 1 \Vdash 1\le \al\\
\rrule{16} & \nil \Vdash 1\le 1\\
\rrule{4}  & \nil
\end{aligned}
\end{gather*}
\caption{A Sample Derivation for Algorithmic Typing}
\label{fig:alg:sample}
\end{figure}

>>>>

\framebox{$\text{wf }\Gm$} Well-formed algorithmic worklist
\begin{gather*}
\inferrule*[right=$\mathtt{wf\nil}$]
{~}{\text{wf }\nil}
\qquad
\inferrule*[right=$\mathtt{wf_a}$]
{\text{wf }\Gm}
{\text{wf }\Gm, a}
\qquad
\inferrule*[right=$\mathtt{wf_{\al}}$]
{\text{wf }\Gm}
{\text{wf }\Gm, \al}
\qquad
\inferrule*[right=$\mathtt{wf_{of}}$]
{\text{wf }\Gm \\ \Gm\vdash A}
{\text{wf }\Gm, x:A}
\qquad
\inferrule*[right=$\mathtt{wf_{\jg}}$]
{\text{wf }\Gm \\ \Gm\vdash\jg}
{\text{wf }\Gm \Vdash\jg}
\end{gather*}
\Description{Extended Syntax and Well-Formedness for the Algorithmic System}
\caption{Extended Syntax and Well-Formedness for the Algorithmic System}\label{fig:alg:syntax}
\end{figure}

@@

% TODO rules too long
\begin{figure}[htp]
\hfill \framebox{$\Gm\rto \Gm'$} \hfill $\Gm$ reduces to $\Gm'$.
%\jimmy{Whenever there is a change on judgment count, marker should be properly added/removed. Decidability: all valid judgment chain should be reduced to nil.}
\begin{gather*}
\begin{aligned}
\Gm, a &\algrule \Gm \qquad
\Gm, \al \algrule \Gm \qquad
\Gm, x:A \algrule \Gm
\\[3mm]
\Gm \Vdash 1\le 1 &\algrule \Gm\\
\Gm \Vdash a\le a &\algrule \Gm\\
\Gm \Vdash \al\le \al &\algrule \Gm\\
\Gm \Vdash A_1\to A_2 \le B_1\to B_2 &\algrule \Gm \Vdash A_2 \le B_2 \Vdash B_1\le A_1\\
\Gm \Vdash \all A\le B &\algrule \Gm,\al \Vdash [\al/a]A\le B \quad\text{when } B \neq \all B'\\
\Gm \Vdash A\le \all[b]B &\algrule \Gm,b \Vdash A\le B
\\[3mm]
%\\
%\text{Let } \color{red}\Gm[\al \toto B // G_M] &:= \Gm_L, G_M, [B/\al]\Gm_R, \text{ when } \Gm[\al] = \Gm_L,\al,\Gm_R \ (G_M\text{ defaults to }\nil)\\
\Gm[\al] \Vdash \al \le A\to B &\algrule [\al[1]\to\al[2]/\al] (\Gm[\al[1], \al[2]] \Vdash \al[1]\to \al[2] \le A \to B)\\
 &\qquad\qquad \text{when }\al\notin FV(A)\cup FV(B)\\
\Gm[\al] \Vdash A\to B \le \al &\algrule [\al[1]\to \al[2]/\al] (\Gm[\al[1], \al[2]] \Vdash A \to B \le \al[1]\to \al[2])\\
 &\qquad\qquad \text{when }\al\notin FV(A)\cup FV(B)
 \\[3mm]
\Gm[\al][\bt] \Vdash \al \le \bt &\algrule [\al/\bt](\Gm[\al][])\\
\Gm[\al][\bt] \Vdash \bt \le \al &\algrule [\al/\bt](\Gm[\al][])\\
\Gm[a][\bt] \Vdash a \le \bt &\algrule [a/\bt](\Gm[a][])\\
\Gm[a][\bt] \Vdash \bt \le a &\algrule [a/\bt](\Gm[a][])\\
\Gm[\bt] \Vdash 1 \le \bt &\algrule [1/\bt](\Gm[])\\
\Gm[\bt] \Vdash \bt \le 1 &\algrule [1/\bt](\Gm[])
\\[3mm]
\Gm \Vdash e \Lto B &\algrule \Gm \Vdash e\To_a a\le B \quad
    \text{when } e \neq \lam e' \text{ and } B \neq \all B'\\
% \Gm \Vdash () \Lto 1 &\algrule \Gm\\
\Gm \Vdash e\Lto \all A &\algrule \Gm,a \Vdash e\Lto A\\
\Gm \Vdash \lam e \Lto A\to B &\algrule \Gm, x:A  \Vdash e \Lto B\\
\Gm[\al] \Vdash \lam e \Lto \al &\algrule [\al[1]\to \al[2] / \al](\Gm[\al[1],\al[2]], x:\al[1] \Vdash e \Lto \al[2])
% \quad\text{\jimmy{Additional}}
\\[3mm]
\Gm \Vdash x\To_a \jg &\algrule \Gm \Vdash [A/a] \jg \quad \text{when } (x:A)\in \Gm\\
\Gm \Vdash (e:A)\To_a \jg &\algrule \Gm \Vdash [A/a]\jg \Vdash e \Lto A\\
\Gm \Vdash ()\To_a \jg &\algrule \Gm \Vdash [1/a]\jg\\
\Gm \Vdash \lam e \To_a \jg &\algrule
    \Gm,\al,\bt \Vdash [\al\to\bt/a]\jg, x:\al \Vdash e\Lto \bt\\
\Gm \Vdash e_1\ e_2 \To_a \jg &\algrule \Gm \Vdash e_1\To_b (\appInfAlg{b}{e_2})
\\[3mm]
\Gm \Vdash \appInfAlg{\all A}{e} &\algrule \Gm,\al \Vdash \appInfAlg{[\al/a]A}{e}\\
\Gm \Vdash \appInfAlg{A\to C}{e} &\algrule \Gm \Vdash [C/a]\jg \Vdash e \Lto A\\
\Gm[\al] \Vdash \appInfAlg{\al}{e} &\algrule
    [\al[1]\to\al[2]/\al](\Gm[\al[1], \al[2]] \Vdash \appInfAlg{\al[1]\to\al[2]}{e})
%	[\al[1]\to\al[2]/\al](\Gm[\al[1], \al[2]] \Vdash [\al[2]/a]\jg \Vdash e\Lto \al[1])\\
% &\color{magenta} \makebox[0pt]{\qquad or} \phantom{{}\rrule{}{}}
\end{aligned}
\end{gather*}
\Description{Algorithmic Typing}
\caption{Algorithmic Typing}\label{fig:alg}
\end{figure}

@@

\paragraph{{\bf 1. Algorithmic subtyping}}
We have six subtyping rules (Rules 4-9) that are similar to their
declarative counterparts. For instance, Rule 7 consumes a subtyping
judgment and pushes two back to the worklist.  Rule 8 differs from
declarative Rule $\mathtt{{\le}{\forall}L}$ by introducing an existential
variable $\al$ instead of guessing the monotype $\tau$
instantiation. Each existential variable is later solved to a
monotype $\tau$ with the same context, so the final solution $\tau$ of
$\al$ should be well-formed under $\Gm$.

@@

\begin{figure}
\begin{gather*}
\begin{aligned}
           & \nil \Vdash (\lam x)~() \Lto 1\\
\rrule{18} & \nil \Vdash (\lam x)~() \To_a a\le 1\\
\rrule{26} & \nil \Vdash (\lam x) \To_b (\appInfAlg{b}{()}[a][a\le 1])\\
\rrule{25} & \nil,\al,\bt \Vdash \appInfAlg{\al\to\bt}{()}[a][a\le 1], x:\al \Vdash x\Lto \bt\\
\rrule{18} & \nil,\al,\bt \Vdash \appInfAlg{\al\to\bt}{()}[a][a\le 1], x:\al \Vdash x\To_b b\le \bt\\
\rrule{22} & \nil,\al,\bt \Vdash \appInfAlg{\al\to\bt}{()}[a][a\le 1], x:\al \Vdash \al\le \bt\\
\rrule{12} & \nil,\al \Vdash \appInfAlg{\al\to\al}{()}[a][a\le 1], x:\al\\
\rrule{3}  & \nil,\al \Vdash \appInfAlg{\al\to\al}{()}[a][a\le 1]\\
\rrule{28} & \nil,\al \Vdash \al\le 1 \Vdash () \Lto \al\\
\rrule{18} & \nil,\al \Vdash \al\le 1 \Vdash () \To_a a\le \al\\
\rrule{24} & \nil,\al \Vdash \al\le 1 \Vdash 1\le \al\\
\rrule{16} & \nil \Vdash 1\le 1\\
\rrule{4}  & \nil
\end{aligned}
\end{gather*}
\Description{A Sample Derivation for Algorithmic Typing}
\caption{A Sample Derivation for Algorithmic Typing}
\label{fig:alg:sample}
\end{figure}
